---
title: "Stat 332 A2 Inventory Analysis"
output: pdf_document
author: "MingMing Z. (21058539)"
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
```

(a) To calculate the 95\% confidence interval for the total error, we follow $\hat{E} \pm c * se(\hat{E})$ where c = 1.96, and $\hat{E}$ is the estimated total error.

```{r}
# (a)
# uses library readr to import data from txt file
df <- read_delim("/Users/ming/Desktop/School/STAT 332/A2/inventory.txt", delim = "\t")
# checking the data frame
head(df)
# separating error values of each strata
erra <- df[df$stratum == "A", "error"]
errb <- df[df$stratum == "B", "error"]
errc <- df[df$stratum == "C", "error"]
# calculating estimated total errors of the separate stratums
tea <- 27 * mean(erra$error, na.rm = TRUE)
teb <- 44 * mean(errb$error, na.rm = TRUE)
tec <- 29 * mean(errc$error, na.rm = TRUE)
# estimated total error:
ehat <- tea + teb + tec
ehat
# error variance for each stratum:
vara <- var(erra$error, na.rm = TRUE)
varb <- var(errb$error, na.rm = TRUE)
varc <- var(errc$error, na.rm = TRUE)
# calculating the total error variance using population and sample proportions as weight
vart <- (27/27) * vara + (2132/44) * varb + (1450/29) * varc
se <- sqrt(vart) #se(ehat)
se

ehat + 1.96 * se
ehat - 1.96 * se
```
So overall, we have [-2159.781, -893.6194] as the 95\% confidence interval for the total error.

(b) We are using proportional allocation for B and C based on the population, which might not be the best if the variance in errors differs a lot between strata B and C. 
```{r}
vara
varb
varc
```
As we can see from the calculation, strata B has a way bigger variance than strata C, so we might consider a bigger sample for B to account for the bigger variance. An optimal allocation approach might be desirable.

(c) We could create more stratas for the data points in B since the variability is quite high. C can remain a stand alone strata since the variance is relatively low compared to the other stratas. A should remain a stand alone strata even though the variance is high since there are only 27 in the population. So a different stratification would just be splitting B into more stratas.
