---
title: "Assignment 3 - Question 3"
author: "MingMing Z (21058539)"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

```{r}
# set up
load("crabs.Rda")
head(crabs)
```

## Part a:

### Part i:

```{r}
# for males (si = 1): \mu(s_i) = theta_1
# for females (si = 0): \mu(s_i) = theta_2

fit <- lm(length ~ factor(sex), data = crabs)
mod <- model.matrix(fit)
head(mod, 3)
# output seems to align with dataset!
# factor(sex)male seems to estimate theta_2-theta_1 when = 1, and just theta_1 when = 0
```
### Part ii:
```{r}
summary(fit)
```
Explain why these values must be identical for this model:
We know $\tilde{\theta} = (X^T X)^{-1} X^T Y \sim N_p(\theta, \sigma^2 (X^T X)^{-1})$, so our standard error will rely on the model matrix X. Our model matrix has an intercept and one indicator variable, and $(X^T X)^{-1}$ is diagonal, leading to similar standard errors for the coefficients.

Explain why the two coefficient estimators must be independently distributed for this model.
Under assumptions of the mean response model, the distributions of all estimators arise from the single stochastic assumption that $Y \sim N_n(\mu, \sigma^2I_n) \implies R \sim N_n(\mu, \sigma^2I_n)$. $\tilde{\theta} \text{ and } \tilde{\mu}$ are distributed independently of both$\tilde{r} \text{ and } \tilde{\sigma}^2$ Therefore the two coefficient estimators must also be independently distributed for this model. More intuitively, these coefficients come from separate sex groups with no overlap, so they must be independent.


### Part iii:

$H_0: \theta_1 = \theta_2$ or $H_0: \theta_2 - \theta_1 = 0$. This is basically hypothesising that factor(sex)male = 0, which we do have the p value from the summary output, but also showing the calculations:
Discrepancy measure: $$d =\frac{(\hat{\theta}_2 - \hat{\theta}_1) - 0}{SE(\hat{\theta}_2 - \hat{\theta}_1)} = \frac{1.4910}{1.0037} \approx 1.485 $$
Using the t distribution with n-p = 200-2 = 198 degrees of freedom we have:
```{r}
2 * pt(-abs(1.485), df = 198)
```
So the p value for this hypothesis is is approximately 0.139, which is >0.1, meaning there is no significant evidence against the null hypothesis. Therefore we can conclude that there is no evidence against the hypothesis that carapace length is the same for both sexes.

## Part b:

### Part i: 
Model 4 has the parameter vector $\theta^T = (\theta_1, \theta_2, \theta_3, \theta_4)$, where
$\theta_1 = \alpha_0$ = sexfemale (intercept parameter for females)
$\theta_2 = \beta_0$ = sexmale (intercept parameter for males)
$\theta_3 = \alpha_1$ = sexfemale:width (slope parameter females)
$\theta_4 = \beta_1$ = sexmale:width (slope parameter males)
We also don't see any $\beta_0$ (Intercept).

### Part ii:

```{r}
#1
# summary(lm(length ~ sex + width, data = crabs))
#2
# summary(lm(length ~ 0 + sex + width, data = crabs))
#3
summary(lm(length ~ sex + sex:width, data = crabs))
#4
summary(lm(length~ 0 + sex + sex:width, data = crabs))
#5
# summary(lm(length~ 0 + sex:width, data = crabs))
#6
summary(lm(length ~ sex*width, data = crabs))
```
We can see here that model 3 and model 6 produces the same residuals and residual standard error as model 4, so would produce the same fit.

### Part iii. 

Model 3:
$\theta_1 = \alpha_0$ = (intercept) (intercept parameter for females)
$\theta_2 = \beta_0 - \alpha_0$ = sexmale (difference in intercets between males and females)
$\theta_3 = \alpha_1$ = sexfemale:width (slope parameter females)
$\theta_4 = \beta_1$ = sexmale:width (slope parameter males)

$$A = \begin{bmatrix}
1 & 0 & 0 & 0 \\
-1 & 0 & 1 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix}$$

Model 6:
$\theta_1 = \alpha_0$ = (intercept) (intercept parameter for females)
$\theta_2 = \beta_0 - \alpha_0$ = sexmale (difference in intercets between males and females)
$\theta_3 = \alpha_1$ = width (slope parameter females)
$\theta_4 = \beta_1 - \alpha_1$ = sexmale:width (difference in slope parameter between males and females)

$$A = \begin{bmatrix}
1 & 0 & 0 & 0 \\
-1 & 0 & 1 & 0 \\
0 & 1 & 0 & 0 \\
0 & -1 & 0 & 1
\end{bmatrix}$$

## Part c:
```{r}
#1
summary(lm(length ~ sex + width, data = crabs))
```
Model 1:
$H_0: \alpha_0 = 0$ - p val = 0.00546 < 0.05, strong evidence against null hypothesis
$H_0: \beta_0 = 0$ N/A
$H_0: \alpha_0 = \beta_0 = \beta_0 = \alpha_1 = \beta_1 = 0$ - p val < 2.2e-16, strong evidence against null hypothesis
```{r}
#2
summary(lm(length ~ 0 + sex + width, data = crabs))
```
Model 2:
$H_0: \alpha_0 = 0$ - p val = 0.00546 < 0.05, strong evidence against null hypothesis
$H_0: \beta_0 = 0$  p val = 0.124 > 0.1, no evidence against null hypothesis
$H_0: \alpha_0 = \beta_0 = \beta_0 = \alpha_1 = \beta_1 = 0$ - p val < 2.2e-16, strong evidence against null hypothesis

```{r}
#3
summary(lm(length ~ sex + sex:width, data = crabs))
```
Model 3:
$H_0: \alpha_0 = 0$ - p val = 0.00188 < 0.05, strong evidence against null hypothesis
$H_0: \beta_0 = 0$ N/A
$H_0: \alpha_0 = \beta_0 = \beta_0 = \alpha_1 = \beta_1 = 0$ - p val < 2.2e-16, strong evidence against null hypothesis

```{r}
#4
fit4<-lm(length~ 0 + sex + sex:width, data = crabs)
summary(fit4)
```
Model 4:
$H_0: \alpha_0 = 0$ - p val = 0.00188 < 0.05, strong evidence against null hypothesis
$H_0: \beta_0 = 0$  p val = 0.6139 > 0.1, no evidence against null hypothesis
$H_0: \alpha_0 = \beta_0 = \beta_0 = \alpha_1 = \beta_1 = 0$ - p val < 2.2e-16, strong evidence against null hypothesis

```{r}
#5
summary(lm(length~ 0 + sex:width, data = crabs))
```
Model 5:
$H_0: \alpha_0 = 0$ N/A
$H_0: \beta_0 = 0$ N/A
$H_0: \alpha_0 = \beta_0 = \beta_0 = \alpha_1 = \beta_1 = 0$ - p val < 2.2e-16, strong evidence against null hypothesis

```{r}
#6
summary(lm(length ~ sex*width, data = crabs))
```
Model 6:
$H_0: \alpha_0 = 0$ - p val = 0.00188 < 0.05, strong evidence against null hypothesis
$H_0: \beta_0 = 0$ N/A
$H_0: \alpha_0 = \beta_0 = \beta_0 = \alpha_1 = \beta_1 = 0$ - p val < 2.2e-16, strong evidence against null hypothesis

Overall:
$H_0: \alpha_0 = 0$ - strong evidence against null hypothesis
$H_0: \beta_0 = 0$ - no evidence against null hypothesis
$H_0: \alpha_0 = \beta_0 = \beta_0 = \alpha_1 = \beta_1 = 0$ - strong evidence against null hypothesis

## Part d.

### Part i.
Using the parameterized mean model, $\alpha_0 + \alpha_1x_i$, we find estimated mean carapace length for female crabs with a maximum carapace width of 40mm to be $$\hat{\alpha_0} + \hat{\alpha_1}(40) = -1.063535 + 0.904927(40) \approx 35.134$$ from model 4 fit results.

To construct a 90% confidence interval, we can use the predict function (since it's only restricted for iii)
We have the 90% confidence interval [35.00479, 35.2623]
``` {r}
predict(fit4, newdata = data.frame(sex = "female", width = 40), interval = "confidence", level = 0.90)
```

### Part ii.
To construct a 90% prediction interval, we can use the predict function (since it's only restricted for iii)
We have the 90% prediction interval [34.00652, 36.26057]
``` {r}
predict(fit4, newdata = data.frame(sex = "female", width = 40), interval = "prediction", level = 0.90)
```

### Part iii.
For a female crab of predicted width 35 mm, we have $-1.063535+0.904927(35)\approx30.609$
For a male crab, we have $\hat{\beta_0} + \hat{\beta_1}(35) = -0.156608+0.892122(35)\approx31.0677$
The difference is approximatly $31.0677 - 30.609 = 0.4587mm$